Bayesian multi-proxy reconstruction of early Eocene latitudinal temperature gradients

### 2023-06-01 manuscript/main.qmd
Update main.qmd
manuscript and cover letter updates

---
title: "Bayesian multi-proxy reconstruction of early Eocene latitudinal temperature gradients"
author:
  - "---"
  - Kilian Eichenseer^1^ and Lewis A. Jones^2^
  - "---"
  - "^1^Department of Earth Sciences, Durham University, South Road, DH1 3LE, Durham, United Kingdom"
  - "^2^Centro de Investigación Mariña, Grupo de Ecoloxía Animal, Departamento de Ecoloxía e Bioloxía Animal, Universidade de Vigo, 36310 Vigo, Spain."
  - "---"
  - "**Corresponding author:** kilian.eichenseer@durham.ac.uk"
abstract: \newpage Abstract
format:
  docx:
    reference-doc: template.docx
    highlight-style: github
    fig-dpi: 900
  pdf:
    documentclass: scrartcl
    papersize: letter
    geometry:
      - top=25.4mm
      - bottom=25.4mm
      - left=25.4mm
      - right=25.4mm
      - heightrounded
header-includes:
  - \usepackage{newunicodechar}
  - \newcommand{\$^\circ$C}{\textdegree C}
bibliography: references.bib
csl: climate-of-the-past.csl
editor: source
---

@@

::: {#table-1}
![](../figures/Table1.png){width="16.25cm"}
Table 1: Inferred latitudinal sea surface temperature (SST) gradients for the early Eocene (EE) or the EECO, as shown in earlier, proxy-based studies. For comparison, a gradient derived from an atmosphere-ocean general circulation model ensemble is also shown.

:::

@@

**Coral reefs.** Today, shallow warm-water coral reefs are limited to tropical and subtropical latitudes (\~34$^\circ$ N -- 32$^\circ$ S), with minimum sea surface temperature tolerances (\~18$^\circ$C) being the primary constraint on this distribution [@johannes1983; @kleypas1999; @yamano2001]. As coral reefs reside at the upper thermal limit of the oceans today, their maximum sea surface temperature tolerance is less well-constrained, with some studies suggesting up to 35.6$^\circ$C in the geological past [@jones2022reef]. Nevertheless, coral reefs have frequently been recognised as tracers of past (sub-)tropical conditions [@ziegler1984; @kiessling2001]. During the Eocene, coral communities and reefs expanded across tropical and temperate latitudes, with communities found up to palaeolatitudes of 43 $^\circ$ N [@Zamagni2012]. Using a compilation of Paleocene--early Eocene coral reefs and community localities [@Zamagni2012], we generated quantitative sea surface temperature estimates for the ECCO. To do so, we extracted localities from the compilation that are inferred to be Ilerdian (early Eocene) coral reefs, and that could be confidently assigned to the EECO. We excluded coral knobs and coral-bearing mounds which might have broader climatic limits than warm-water coral reef ecosystems. This filtering resulted in four unique coral reef localities remaining for the EECO, all of which conform to the modern latitudinal range of coral reefs (\<34$^\circ$ N). Subsequently, we used statistically derived temperature limits (minimum = 21$^\circ$C, average = 27.6$^\circ$C, maximum = 29.5$^\circ$C) from the published literature [@kleypas1999] to define a normal probability distribution of potential temperature values for coral reef localities. This normal probability distribution was defined with a mean of 27.6 and a standard deviation of 2.125, placing 97.5% of the probability density above the minimum. As the distribution of modern corals is skewed towards warmer temperatures, this approach results in 16.5% of the probability being placed on temperatures \> 29.5$^\circ$C, allowing for the possibility that Eocene coral reefs were adapted to warmer conditions than present-day coral reefs.

**Mangroves.** Mangroves are distributed throughout the tropics and subtropics today. While factors besides sea surface temperatures (SST) influence the distribution of mangroves, empirical, lower temperature limits have been established for the genera *Avicennia* (15.6$^\circ$C) and *Rhizophora* (20.7$^\circ$C) [@Quisthoudt2012]. Both *Avicennia* and members of the Rhizophoraceae family were widespread and co-occurred across tropical and temperate latitudes in the early Eocene. Only *Avicennia*, however, occurred at polar latitudes [@Suan2017; @Popescu2021]. Assuming that Eocene members of these mangrove taxa conform to similar climatic requirements as their modern relatives, the presence and absence of *Avicennia* and Rhizophoraceae pollen can be used as a palaeotemperature indicator. For this analysis, published mangrove occurrence data were taken from @Popescu2021, and converted to quantitative temperature estimates. From this data, we identify two types of pollen assemblages which we ascribe different temperature distributions:

1)  *Avicennia*-only assemblages ($n=2$): the absence of Rhizophoraceae is indicative of temperatures being between 15.6$^\circ$C (lower temperature limt of *Avicennia*) and 20.7$^\circ$C (lower temperature limit of *Rhizophora*). However, a value of 22.5$^\circ$C is ascribed as the upper temperature limit here as *Rhizophora* is rare below this temperature. We define the *Avicennia*-only temperature distribution as a normal distribution with a mean of 19.05 and a standard deviation of 1.725, resulting in 95% of the probability density being placed within the temperature limits.

2)  *Avicennia* and Rhizophoraceae assemblages ($n=5$): the presence of both groups suggests that the locality should have a minimum temperature of 20.7$^\circ$C (lower temperature limit of *Rhizophora*). As the upper thermal limits of *Aviciennia* and *Rhizophora* are not well established in @Quisthoudt2012, we assign the same maximum temperature limits (29.5$^\circ$C) as coral reef localities, because mangroves are also widely distributed throughout tropical regions. Consequently, we define the temperature distribution for this locality as a normal distribution with a mean of 25.1 and a standard deviation of 2.2, with 95% probability density within the temperature limits.

@@

To test whether our logistic regression model can adequately describe different latitudinal temperature gradients at various sample sizes, we generated four idealised gradients that emulate potential climatic states throughout Earth's geological history: extreme icehouse, icehouse, greenhouse, and extreme greenhouse [@Frakes1992]. We then randomly sampled (1000 iterations) these gradients using increasing sample sizes (5, 10, and 20) and reconstructed the latitudinal temperature gradient using our model for each of these sample sizes and gradient types. Using the same idealised gradients, we also tested whether our model could accurately reconstruct latitudinal temperature gradients using the palaeogeographic distribution of Eocene samples (*n* = 34), providing an empirical, exemplary distribution that captures both limited sample size and skewed geographic origins of samples. To evaluate how well the model performed in reconstructing the idealised gradients from limited sampling, we calculated the coefficient of determination ($R^2$) for Bayesian regression models [@Gelman2019]. For every iteration from the posterior, we intercepted the modelled and the idealised gradient in intervals of 1$^\circ$ latitude and calculated the $R^2$ based on these values. We report the median, and 95% credible intervals (CI) of the resulting $R^2$ values. Here and in all other instances, the 95% CI refer to the interval between the 2.5% point and the 97.5% point of the samples or sampled posterior distribution.

To test whether our model can accurately depict the shape of the modern sea surface temperature gradient, and to facilitate comparison with the Eocene gradient, we applied our model to annual sea surface mean temperatures from Bio-Oracle [@Assis2018], aggregated to a $1^\circ \times 1 ^\circ$ raster (*n* = 46,131). The $R^2$ for the modern gradient was calculated as above [@Gelman2019], comparing the modelled gradient and the empirical temperature averages in 1$^\circ$ latitude bins. Only the medians are reported for the modern gradient, as the 95% credible intervals are extremely narrow due to the high precision of the posterior estimates.

@@

The latitudinal gradient is calculated as the difference between the modelled temperature at the equator (0$^\circ$ latitude) and at the poles (90$^\circ$ absolute latitude). To facilitate comparison with earlier estimates, we also calculate the gradient with the temperature at the polar circle (66.6$^\circ$ absolute latitude) being used instead of at the poles. Given the sigmoidal shape of the modern as well as the Eocene gradient (see Fig. 4), these results are broadly comparable to a gradient inferred from the zonal average of equatorial and high-latitude temperatures, as is done in some earlier studies [@Evans2018].

@@

Global average temperatures with 95% credible intervals were calculated by taking the weighted mean of the median (95% CI) of temperature estimates in 1$^\circ$ latitudinal bins. The weights were set to the proportion of global surface area in each latitudinal bin, i.e. decreasing with increasing latitude as: $$weights = sin(\alpha_{1,i}) - sin(\alpha_{2,i}) ,$${#eq-weights} where $\alpha_1$ is the upper, and $\alpha_2$ is the lower latitudinal boundary of bin $i$, i.e. we approximated the shape of the globe as a spheroid.

@@

The average, modern temperature gradient can be closely approximated with our model when using the full modern SST dataset (Fig. 3); almost all of the variation in the empirical median temperatures in bins of 1$^\circ$ absolute latitude is explained by the modelled gradient ($R^2 = `r R2_modern`$). The empirical gradient spans `r modern_gradient_empirical`$^\circ$C from the equator to the poles, the modelled gradient is only very slightly higher at `r modern_gradient$"50%"`$^\circ$C. The modern, global mean temperature (GMST) based on our modelled, median gradient is `r motemp_global$"50%"`$^\circ$C, very similar to the GMST derived from the empirical median gradient (`r motemp_empirical_global`$^\circ$C).

<!-- REMOVED
Reducing the amount and the spatial coverage of data points to match the early Eocene sampling distribution ($n = 34$) still results in a close match of the modelled gradient with the empirical gradient (Fig. 3b). The 95% credible interval (CI) of reconstructed gradients (Fig. 3b) is noticeably narrower than the spread of empirical temperature values (Fig. 3a). The modern, latitudinal gradient, modelled with all data, spans `r modern_gradient$"50%"`$^\circ$C from the equator to the poles. When modelled with only the early Eocene sampling locations, the median temperature gradient is `r modern_eocene_gradient$"50%"` (95% CI: `r modern_eocene_gradient$"2.5%"` -- `r modern_eocene_gradient$"97.5%"`)$^\circ$C. A random sampling of modern temperatures at latitudes corresponding to the sampled Eocene palaeolatitudes would thus allow the reconstruction of the modern latitudinal temperature gradient with good accuracy, with expected deviations from the actual gradient as low as `r modern_eocene_gradient_diff$"2.5%"` to `r modern_eocene_gradient_diff$"97.5%"`$^\circ$C (95 % CI).

@@

modelling the latitudinal temperature gradient facilitates accurate reconstruction of global mean surface temperatures (GMST). The modern, global mean sea surface temperature (GMsST) estimate is `r motemp_global$"50%"`$^\circ$C when using the full, modern data set. When reduced to the early Eocene sampling distribution, the modern GMsST is estimated at `r moeotemp_global$"50%"` (`r moeotemp_global$"2.5%"` -- `r moeotemp_global$"97.5%"`)$^\circ$C. The deviations from the full modern estimate are again very low, with a 95 % credible interval of `r moeotemp_global_diff$"2.5%"` to `r moeotemp_global_diff$"97.5%"`$^\circ$C.
-->

@@

The modelled Eocene temperature gradient is starkly different from the modern (Fig 4). Modelled, median equatorial temperatures are `r tdiff_lat0$"50%"` (95% CI: `r tdiff_lat0$"2.5%"` -- `r tdiff_lat0$"97.5%"`)$^\circ$C higher for the EECO, and polar temperatures are `r tdiff_lat90$"50%"` (`r tdiff_lat90$"2.5%"` -- `r tdiff_lat90$"97.5%"`)$^\circ$C higher. This results in a flattened latitudinal temperature gradient of `r eocene_gradient$"50%"` (`r eocene_gradient$"2.5%"` -- `r eocene_gradient$"97.5%"`)$^\circ$C for the EECO, as opposed to `r modern_gradient$"50%"`$^\circ$C for the modern. To facilitate the comparison with latitudinal gradients reported in the literature, which sometimes do not report temperatures at very high latitudes, we report also the EECO gradient between the equator and the modern-day polar circle (66.6$^\circ$), which is slightly lower at `r eocene_gradient_polar_circle$"50%"` (`r eocene_gradient_polar_circle$"2.5%"` -- `r eocene_gradient_polar_circle$"97.5%"`)$^\circ$C.

The high variability of EECO palaeotemperature proxies, particularly in the mid-latitudes, and the scarcity of high-latitude data, result in substantial uncertainties in the modelled temperature gradient. This is reflected in the residual standard deviation ($\sigma$) of the EECO gradient -- `r eeco_sdy$"50%"` (`r eeco_sdy$"2.5%"` -- `r eeco_sdy$"97.5%"`)$^\circ$C -- which is more than double the  $\sigma$ for the modern gradient, `r modern_sdy$"50%"`. This signifies that the early Eocene data does not fit as well to the logistic latitudinal gradient model, which can also be seen from the drastic departure of some of the proxy data from the gradient estimates (Fig. 4).

The early Eocene GMST is estimated at `r eotemp_global$"50%"` (`r eotemp_global$"2.5%"` -- `r eotemp_global$"97.5%"`)$^\circ$C, `r  round(as.numeric(eotemp_global$"50%") - as.numeric(motemp_global$"50%"),1)`$^\circ$C higher than the modern. A model run excluding the ecological proxies increases the GMST by `r eoceneproxy_global_diff$"50%"` (`r eoceneproxy_global_diff$"2.5%"` -- `r eoceneproxy_global_diff$"97.5%"`)$^\circ$C. The median latitudinal gradient is similar when excluding the ecological proxies, with a median of `r eocene_proxy_gradient$"50%"`$^\circ$C, but with a `r proxy_uncertain_diff_percent`% wider 95% CI (Fig. S2). This indicates that the ecological proxy data are broadly in agreement with the geochemical proxies, while providing additional constraints on the shape of the early Eocene temperature gradient.

Due to the limited spatial coverage of the early Eocene proxy record, and due to the added model complexity of simultaneously estimating a model across both hemispheres, we pooled the proxy data across both hemispheres. Applying the model separately within each hemisphere results in substantial differences in hemispherical, average temperatures, with the Southern Hemisphere being warmer by `r eeco_south_north_global_diff$"50%"` (`r eeco_south_north_global_diff$"2.5%"` -- `r eeco_south_north_global_diff$"97.5%"`)$^\circ$C. The inferred latitudinal gradient is somewhat steeper in the Northern Hemisphere (steeper by `r eocene_north_south_gradient_diff$"50%"`$^\circ$C, although the 95% CI spans `r eocene_north_south_gradient_diff$"2.5%"` -- `r eocene_north_south_gradient_diff$"97.5%"`$^\circ$C), but the large uncertainties associated with both gradients, and the lack of polar proxy data in the Southern Hemisphere preclude a more precise statement (see Fig. S3).

@@

Our results further exemplify how incorporating quantified ecological temperature constraints can provide more precise temperature reconstructions than geochemical proxies alone, adding to the advances in palaeoclimate reconstructions achieved by integrating lithological data [@Scotese2021; @Burgener2023]. Combining the occurrences of climate-sensitive plant communities [@Greenwood1995], reptiles [@Markwick2007], leaf shapes [@Peppe2011], with geochemical proxies offers substantial potential for improving quantitative palaeoclimate reconstructions across the Phanerozoic. Our modelling framework offers a straightforward, efficient way of integrating ecological climate data with other proxy data: The hierarchical model structure accounts for variation of temperature estimates from proxies at individual localities, which is treated equivalent to the uncertainty associated with the ecological temperature proxies. A local temperature estimate, based on multiple geochemical proxies, thus has the same weight as a local temperature estimate obtained from the occurrence of a climate-sensitive plant community, whilst preserving the uncertainty associated with each estimate. The model could easily be extended to include uncertainties on individual geochemical proxy data, or to differently weight proxy records classified as more or less reliable.

Our approach for deriving fully quantitative climate reconstructions from ecological data is borrowed from nearest living relative methods, commonly employed in terrestrial, Cenozoic climate reconstructions [@Fauquette2007; @Pross2012]. One major limitation to these methods is that the thermal preferences of taxa may have changed over time. More significantly, in the early Eocene, sea surface temperatures may have reached heights unknown in the modern world, and nearest living relative methods based on the modern are inherently unable to predict such elevated temperatures. This is especially true for taxa that inhabit the warmest part of the ocean today, e.g. coral reefs [@kleypas1999]. Although coral reefs are threatened by warming sea surface temperatures today [@Hoegh2011], it is conceivable that Eocene reef corals were adapted to a warmer climate. The fossil record indicates that reef development may have been stunted in the early Eocene, with few early Eocene coral reefs occurring in low latitudes [@Zamagni2012]. The absence of coral reefs in higher latitudes in the early Eocene could be due to requirements in irradiance, rather than temperature [@Muir2015]. Tropical temperatures predicted by the geochemical proxy record indicate hotter-than-modern tropical temperatures for the early Eocene (Fig. S2), suggesting that the modern climate range of coral reefs may underestimate the early Eocene thermal niche for coral reefs. We have tried to account for that possibility by widening the temperature probability distribution for reef corals, but the predicted temperatures for the reef and mangrove sites still lie below the temperatures indicated by the geochemical proxy record (Fig. 4, Fig. S2).

@@

Both authors designed the study and carried out data preparation. K.E. programmed the model and conducted the analyses. L.A.J. generated the figures. Both authors contributed to the writing of the manuscript.

>>>>

---
title: "Bayesian multi-proxy reconstruction of early Eocene latitudinal temperature gradients"
author:
  - "---"
  - Kilian Eichenseer^1^ and Lewis A. Jones^2^
  - "---"
  - "^1^Department of Earth Sciences, Durham University, South Road, DH1 3LE, Durham, United Kingdom"
  - "^2^Centro de Investigación Mariña, Grupo de Ecoloxía Animal, Departamento de Ecoloxía e Bioloxía Animal, Universidade de Vigo, 36310 Vigo, Spain."
  - "---"
  - "**Corresponding author:** kilian.eichenseer@durham.ac.uk"
abstract: \newpage Abstract
format:
  docx:
    reference-doc: template.docx
    highlight-style: github
    fig-dpi: 900
  pdf:
    documentclass: scrartcl
    papersize: letter
    geometry:
      - top=25.4mm
      - bottom=25.4mm
      - left=25.4mm
      - right=25.4mm
      - heightrounded
header-includes:
  - \usepackage{newunicodechar}
  - \newcommand{\&deg;C}{\textdegree C}
bibliography: references.bib
csl: climate-of-the-past.csl
editor: source
---

@@

::: {#table-1}
Table 1: Inferred latitudinal sea surface temperature (SST) gradients for the early Eocene (EE) or the EECO, as shown in earlier, proxy-based studies. For comparison, a gradient derived from an atmosphere-ocean general circulation model (GCM) ensemble is also shown.

```{r, results='asis', echo = FALSE}
library(knitr)

data <- data.frame(
  Source = c("Bijl et al. (2009)", "Keating-Bitoni et al. (2011)", "Tierney et al. (2017)", "Cramwinckel et al. (2018)", "Evans et al. (2018)", "Pross et al. (2012), as shown in Tierney et al. (2017)"),
  Time = c("EE", "EECO", "EE", "EECO", "EE", "EE"),
  Gradient = c("7", "13", "12", "21 (±1)", "20 (±3)", "26"),
  Type_of_gradient = c("equator - polar circle", "equator - polar circle", "equator - polar circle", "equator - deep water", "tropics - deep water", "equator - polar circle"),
  Model = c("$2^{nd}$ order polynomial",
            "$2^{nd}$ order polynomial",
            "Gaussian function",
            "-",
            "-",
            "climate model ensemble (GCM)"),
  Proxy_system = c("$TEX_{86}$, $UK_{37}^{K'}$",
                   "$TEX_{86}$, MBT/CBT, $\\Delta_{47}$, Mg/Ca, $\\delta^{18}O$",
                   "$TEX_{86}$",
                   "$TEX_{86}$, $\\Delta_{47}$, Mg/Ca, $\\delta^{18}O$, deepwater $\\delta^{18}O$",
                   "$\\Delta_{47}$, deepwater Mg/Ca",
                   "none (GCM simulations)")
)

kable(data, format = "markdown")
```
:::

@@

**Coral reefs.** Today, shallow warm-water coral reefs are limited to tropical and subtropical latitudes (\~34&deg;&nbsp;N -- 32&deg;&nbsp;S), with minimum sea surface temperature tolerances (\~18&deg;C) being the primary constraint on this distribution [@johannes1983; @kleypas1999; @yamano2001]. As coral reefs reside at the upper thermal limit of the oceans today, their maximum sea surface temperature tolerance is less well-constrained, with some studies suggesting up to 35.6&deg;C in the geological past [@jones2022reef]. Nevertheless, coral reefs have frequently been recognised as tracers of past (sub-)tropical conditions [@ziegler1984; @kiessling2001]. During the Eocene, coral communities and reefs expanded across tropical and temperate latitudes, with communities found up to palaeolatitudes of 43 &deg;&nbsp;N [@Zamagni2012]. Using a compilation of Paleocene--early Eocene coral reefs and community localities [@Zamagni2012], we generated quantitative sea surface temperature estimates for the ECCO. To do so, we extracted localities from the compilation that are inferred to be Ilerdian (early Eocene) coral reefs, and that could be confidently assigned to the EECO. We excluded coral knobs and coral-bearing mounds which might have broader climatic limits than warm-water coral reef ecosystems. This filtering resulted in four unique coral reef localities remaining for the EECO, all of which conform to the modern latitudinal range of coral reefs (\<34&deg;&nbsp;N). Subsequently, we used statistically derived temperature limits (minimum = 21&deg;C, average = 27.6&deg;C, maximum = 29.5&deg;C) from the published literature [@kleypas1999] to define a normal probability distribution of potential temperature values for coral reef localities. This normal probability distribution was defined with a mean of 27.6 and a standard deviation of 2.125, placing 97.5% of the probability density above the minimum. As the distribution of modern corals is skewed towards warmer temperatures, this approach results in 16.5% of the probability being placed on temperatures \> 29.5&deg;C, allowing for the possibility that Eocene coral reefs were adapted to warmer conditions than present-day coral reefs.

**Mangroves.** Mangroves are distributed throughout the tropics and subtropics today. While factors besides sea surface temperatures (SST) influence the distribution of mangroves, empirical, lower temperature limits have been established for the genera *Avicennia* (15.6&deg;C) and *Rhizophora* (20.7&deg;C) [@Quisthoudt2012]. Both *Avicennia* and members of the Rhizophoraceae family were widespread and co-occurred across tropical and temperate latitudes in the early Eocene. Only *Avicennia*, however, occurred at polar latitudes [@Suan2017; @Popescu2021]. Assuming that Eocene members of these mangrove taxa conform to similar climatic requirements as their modern relatives, the presence and absence of *Avicennia* and Rhizophoraceae pollen can be used as a palaeotemperature indicator. For this analysis, published mangrove occurrence data were taken from @Popescu2021, and converted to quantitative temperature estimates. From this data, we identify two types of pollen assemblages which we ascribe different temperature distributions:

1)  *Avicennia*-only assemblages ($n=2$): the absence of Rhizophoraceae is indicative of temperatures being between 15.6&deg;C (lower temperature limt of *Avicennia*) and 20.7&deg;C (lower temperature limit of *Rhizophora*). However, a value of 22.5&deg;C is ascribed as the upper temperature limit here as *Rhizophora* is rare below this temperature. We define the *Avicennia*-only temperature distribution as a normal distribution with a mean of 19.05 and a standard deviation of 1.725, resulting in 95% of the probability density being placed within the temperature limits.

2)  *Avicennia* and Rhizophoraceae assemblages ($n=5$): the presence of both groups suggests that the locality should have a minimum temperature of 20.7&deg;C (lower temperature limit of *Rhizophora*). As the upper thermal limits of *Aviciennia* and *Rhizophora* are not well established in @Quisthoudt2012, we assign the same maximum temperature limits (29.5&deg;C) as coral reef localities, because mangroves are also widely distributed throughout tropical regions. Consequently, we define the temperature distribution for this locality as a normal distribution with a mean of 25.1 and a standard deviation of 2.2, with 95% probability density within the temperature limits.

@@

To test whether our logistic regression model can adequately describe different latitudinal temperature gradients at various sample sizes, we generated four idealised gradients that emulate potential climatic states throughout Earth's geological history: extreme icehouse, icehouse, greenhouse, and extreme greenhouse [@Frakes1992]. We then randomly sampled (1000 iterations) these gradients using increasing sample sizes (5, 10, and 20) and reconstructed the latitudinal temperature gradient using our model for each of these sample sizes and gradient types. Using the same idealised gradients, we also tested whether our model could accurately reconstruct latitudinal temperature gradients using the palaeogeographic distribution of Eocene samples (*n* = 34), providing an empirical, exemplary distribution that captures both limited sample size and skewed geographic origins of samples. To evaluate how well the model performed in reconstructing the idealised gradients from limited sampling, we calculated the coefficient of determination ($R^2$) for Bayesian regression models [@Gelman2019]. For every iteration from the posterior, we intercepted the modelled and the idealised gradient in intervals of 1&deg; latitude and calculated the $R^2$ based on these values. We report the median, and 95% credible intervals (CI) of the resulting $R^2$ values. Here and in all other instances, the 95% CI refer to the interval between the 2.5% point and the 97.5% point of the samples or sampled posterior distribution.

To test whether our model can accurately depict the shape of the modern sea surface temperature gradient, and to facilitate comparison with the Eocene gradient, we applied our model to annual sea surface mean temperatures from Bio-Oracle [@Assis2018], aggregated to a $1^\circ \times 1 ^\circ$ raster (*n* = 46,131). The $R^2$ for the modern gradient was calculated as above [@Gelman2019], comparing the modelled gradient and the empirical temperature averages in 1&deg; latitude bins. Only the medians are reported for the modern gradient, as the 95% credible intervals are extremely narrow due to the high precision of the posterior estimates.

@@

The latitudinal gradient is calculated as the difference between the modelled temperature at the equator (0&deg; latitude) and at the poles (90&deg; absolute latitude). To facilitate comparison with earlier estimates, we also calculate the gradient with the temperature at the polar circle (66.6&deg; absolute latitude) being used instead of at the poles. Given the sigmoidal shape of the modern as well as the Eocene gradient (see Fig. 4), these results are broadly comparable to a gradient inferred from the zonal average of equatorial and high-latitude temperatures, as is done in some earlier studies [@Evans2018].

@@

Global average temperatures with 95% credible intervals were calculated by taking the weighted mean of the median (95% CI) of temperature estimates in 1&deg; latitudinal bins. The weights were set to the proportion of global surface area in each latitudinal bin, i.e. decreasing with increasing latitude as: $$weights = sin(\alpha_{1,i}) - sin(\alpha_{2,i}) ,$${#eq-weights} where $\alpha_1$ is the upper, and $\alpha_2$ is the lower latitudinal boundary of bin $i$, i.e. we approximated the shape of the globe as a spheroid.

@@

The average, modern temperature gradient can be closely approximated with our model when using the full modern SST dataset (Fig. 3); almost all of the variation in the empirical median temperatures in bins of 1&deg; absolute latitude is explained by the modelled gradient ($R^2 = `r R2_modern`$). The empirical gradient spans `r modern_gradient_empirical`&deg;C from the equator to the poles, the modelled gradient is only very slightly higher at `r modern_gradient$"50%"`&deg;C. The modern, global mean temperature (GMST) based on our modelled, median gradient is `r motemp_global$"50%"`&deg;C, very similar to the GMST derived from the empirical median gradient (`r motemp_empirical_global`&deg;C).

<!-- REMOVED
Reducing the amount and the spatial coverage of data points to match the early Eocene sampling distribution ($n = 34$) still results in a close match of the modelled gradient with the empirical gradient (Fig. 3b). The 95% credible interval (CI) of reconstructed gradients (Fig. 3b) is noticeably narrower than the spread of empirical temperature values (Fig. 3a). The modern, latitudinal gradient, modelled with all data, spans `r modern_gradient$"50%"`&deg;C from the equator to the poles. When modelled with only the early Eocene sampling locations, the median temperature gradient is `r modern_eocene_gradient$"50%"` (95% CI: `r modern_eocene_gradient$"2.5%"` -- `r modern_eocene_gradient$"97.5%"`)&deg;C. A random sampling of modern temperatures at latitudes corresponding to the sampled Eocene palaeolatitudes would thus allow the reconstruction of the modern latitudinal temperature gradient with good accuracy, with expected deviations from the actual gradient as low as `r modern_eocene_gradient_diff$"2.5%"` to `r modern_eocene_gradient_diff$"97.5%"`&deg;C (95 % CI).

@@

modelling the latitudinal temperature gradient facilitates accurate reconstruction of global mean surface temperatures (GMST). The modern, global mean sea surface temperature (GMsST) estimate is `r motemp_global$"50%"`&deg;C when using the full, modern data set. When reduced to the early Eocene sampling distribution, the modern GMsST is estimated at `r moeotemp_global$"50%"` (`r moeotemp_global$"2.5%"` -- `r moeotemp_global$"97.5%"`)&deg;C. The deviations from the full modern estimate are again very low, with a 95 % credible interval of `r moeotemp_global_diff$"2.5%"` to `r moeotemp_global_diff$"97.5%"`&deg;C.
-->

@@

The modelled Eocene temperature gradient is starkly different from the modern (Fig 4). Modelled, median equatorial temperatures are `r tdiff_lat0$"50%"` (95% CI: `r tdiff_lat0$"2.5%"` -- `r tdiff_lat0$"97.5%"`)&deg;C higher for the EECO, and polar temperatures are `r tdiff_lat90$"50%"` (`r tdiff_lat90$"2.5%"` -- `r tdiff_lat90$"97.5%"`)&deg;C higher. This results in a flattened latitudinal temperature gradient of `r eocene_gradient$"50%"` (`r eocene_gradient$"2.5%"` -- `r eocene_gradient$"97.5%"`)&deg;C for the EECO, as opposed to `r modern_gradient$"50%"`&deg;C for the modern. To facilitate the comparison with latitudinal gradients reported in the literature, which sometimes do not report temperatures at very high latitudes, we report also the EECO gradient between the equator and the modern-day polar circle (66.6&deg;), which is slightly lower at `r eocene_gradient_polar_circle$"50%"` (`r eocene_gradient_polar_circle$"2.5%"` -- `r eocene_gradient_polar_circle$"97.5%"`)&deg;C.

The high variability of EECO palaeotemperature proxies, particularly in the mid-latitudes, and the scarcity of high-latitude data, result in substantial uncertainties in the modelled temperature gradient. This is reflected in the residual standard deviation ($\sigma$) of the EECO gradient -- `r eeco_sdy$"50%"` (`r eeco_sdy$"2.5%"` -- `r eeco_sdy$"97.5%"`)&deg;C -- which is more than double the  $\sigma$ for the modern gradient, `r modern_sdy$"50%"`. This signifies that the early Eocene data does not fit as well to the logistic latitudinal gradient model, which can also be seen from the drastic departure of some of the proxy data from the gradient estimates (Fig. 4).

The early Eocene GMST is estimated at `r eotemp_global$"50%"` (`r eotemp_global$"2.5%"` -- `r eotemp_global$"97.5%"`)&deg;C, `r  round(as.numeric(eotemp_global$"50%") - as.numeric(motemp_global$"50%"),1)`&deg;C higher than the modern. A model run excluding the ecological proxies increases the GMST by `r eoceneproxy_global_diff$"50%"` (`r eoceneproxy_global_diff$"2.5%"` -- `r eoceneproxy_global_diff$"97.5%"`)&deg;C. The median latitudinal gradient is similar when excluding the ecological proxies, with a median of `r eocene_proxy_gradient$"50%"`&deg;C, but with a `r proxy_uncertain_diff_percent`% wider 95% CI (Fig. S2). This indicates that the ecological proxy data are broadly in agreement with the geochemical proxies, while providing additional constraints on the shape of the early Eocene temperature gradient.

Due to the limited spatial coverage of the early Eocene proxy record, and due to the added model complexity of simultaneously estimating a model across both hemispheres, we pooled the proxy data across both hemispheres. Applying the model separately within each hemisphere results in substantial differences in hemispherical, average temperatures, with the Southern Hemisphere being warmer by `r eeco_south_north_global_diff$"50%"` (`r eeco_south_north_global_diff$"2.5%"` -- `r eeco_south_north_global_diff$"97.5%"`)&deg;C. The inferred latitudinal gradient is somewhat steeper in the Northern Hemisphere (steeper by `r eocene_north_south_gradient_diff$"50%"`&deg;C, although the 95% CI spans `r eocene_north_south_gradient_diff$"2.5%"` -- `r eocene_north_south_gradient_diff$"97.5%"`&deg;C), but the large uncertainties associated with both gradients, and the lack of polar proxy data in the Southern Hemisphere preclude a more precise statement (see Fig. S3).

@@

Our results further exemplify how incorporating quantified ecological temperature constraints can provide more precise temperature reconstructions than geochemical proxies alone, adding to the advances in palaeoclimate reconstructions achieved by integrating lithological data [@Scotese2021; @Burgener2023]. Combining the occurrences of climate-sensitive plant communities [@Greenwood1995], reptiles [@Markwick2007], leaf shapes [@Peppe2011], with geochemical proxies offers substantial potential for improving quantitative palaeoclimate reconstructions across the Phanerozoic. Our modelling framework offers a straightforward, efficient way of integrating ecological climate data with other proxy data: The hierarchical model structure accounts for variation of temperature estimates from proxies at individual localities, which is treated equivalent to the uncertainty associated with the ecological temperature proxies. A local temperature estimate, based on multiple geochemical proxies, thus has the same weight as a local temperature estimate obtained from the occurrence of a climate-sensitive plant community, whilst preserving the uncertainty associated with each estimate. The model could easily be extended to include uncertainties on individual geochemical proxy data, or to variably weight proxy records classified as more or less reliable.

Our approach for deriving fully quantitative climate reconstructions from ecological data is borrowed from nearest living relative methods, commonly employed in terrestrial, Cenozoic climate reconstructions [@Fauquette2007; @Pross2012]. One major limitation to these methods is that the thermal preferences of taxa may have changed over time. More significantly, in the early Eocene, sea surface temperatures may have reached heights unknown in the modern world, and nearest living relative methods based on the modern are inherently unable to predict such elevated temperatures. This is especially true for taxa that inhabit the warmest part of the ocean today, e.g. coral reefs [@kleypas1999]. Although coral reefs are threatened by warming sea surface temperatures today [@Hoegh2011], it is conceivable that Eocene reef corals were adapted to a warmer climate. The fossil record indicates that reef development may have been stunted in the early Eocene, with few early Eocene coral reefs occurring in low latitudes [@Zamagni2012]. The absence of coral reefs in higher latitudes in the early Eocene could be due to requirements in irradiance, rather than temperature [@Muir2015]. Tropical temperatures predicted by the geochemical proxy record indicate hotter-than-modern tropical temperatures for the early Eocene (Fig. S2), suggesting that the modern climate range of coral reefs may underestimate the early Eocene thermal niche for coral reefs. We have tried to account for that possibility by widening the temperature probability distribution for coral reefs, but the predicted temperatures for the reef and mangrove sites still lie below the temperatures indicated by the geochemical proxy record (Fig. 4, Fig. S2).

@@

Both authors designed the study and carried out data preparation. K.E. programmed the model and conducted the analyses. L.A.J. and K.E. generated the figures. Both authors contributed to the writing of the manuscript.
